### 2010-12-07 Documentation/atropos/atropos.tex
Updates to introduction and some FIXMEs
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@689 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\begin{center}
{\Large \bf An Open Source Framework for Multivariate $n$-Tissue Segmentation and Volumetric Cortical Parcellation:  Evaluation Using the BrainWeb, NIREP, and LPBA Public Data Sets} \\
\vspace*{0.5cm}
{\normalsize Brian B.~Avants$^{1*}$,  Nicholas J.~Tustison$^2$%
\symbolfootnote[1]{
  The first two authors contributed equally to this work.
}, 
Philip A.~Cook$^1$, Jue Woo$^1$, and James C.~Gee$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
\end{singlespace}
\end{center}

@@

\begin{abstract} 
Neuroanatomical coordinate systems are essential for the interpretation of 
structural and functional imaging studies.  However, manual delineation of the 
sulco-gyral complex is time consuming and prone to variability given cortical 
complexity.  This work describes an open source, image-based algorithmic approach to 
cortical parcellation which uses training data to propagate structural labelings to 
individual images.  Along with other neuroinformatics tools publicly distributed in our
ANTs (Advanced Normalization Tools)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
package, {\em Atropos} is an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation algorithm which uses either
parametric or non-parametric finite mixture modelings (FMM) of the
class intensities.  It is capable of incorporating prior labelings,
prior probability maps, and/or Markov Random Field (MRF) modeling
assumptions as spatial priors and geodesically propagating
segmentation information across neuroanatomy to ensure a dense
labeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 68 classes) with a minimal memory footprint.  In addition
to describing the technical and implementation aspects of Atropos,
this work evaluates its performance on the parcellation problem given
two different training datasets, the BrainWeb dataset from Montreal
Neurological Institute and the Hammers atlas from UCL [FIXME].  We use
the datasets to evaluate three uses of the Atropos segmentation method: (1) K-means
segmentation without use of template data; (2) segmentation with prior
probability maps derived from a group template; (3) N-class EM
segmentation for parcellating the brain into 68 classes based on
training data.  These evaluation studies, combined with illustrative
examples that excercise Atropos options, demonstrate both performance
and wide applicability of this new segmentation tool.

%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.

\end{abstract}

@@

\section{Introduction} Neurological and neurodevelopmental research
rely on anatomical and functional correspondence both within and
between subjects.  As image acquisition technology has advanced,
concomitant investment has been made towards image classification
techniques for the cerebrum beginning with the seminal work of
\cite{Vannier1985} in which NASA satellite image processing software
was appropriated for statistical classification of head tissues in 2-D
MR images.  Since then, a proliferation of techniques has ensued with
increasing sophistication.  The chronology of progress in segmentation
may be tracked through both technical reviews
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
and evaluation studies
\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

@@

Observations regarding a general trend towards a more integrative view 
of neuroanatomical image processing led to the work described in \cite{Ashburner2005} which is
publicly available as SPM---a large-scale Matlab module in which registration, segmentation, and bias field 
correction can be simultaneously modeled within a single optimization scheme. The roots of this very popular 
software package stem back to early seminal work by Karl Friston in which the basis for statistical parametric mapping was developed \citep{Friston1990} and subsequently provided to the research community. A similar holistic approach
to brain processing was provided in \cite{Pohl2006} in which segmentation
and registration parameters were optimized simultaneously while casting the inhomogeneity model parameters
of \cite{Wells1996} as nuisance variables.  Continued work involved recursive parcellation of the brain volume by considering sub-structures in a hierarchical manner \citep{Pohl2007}.  An implementation is provided in  3D slicer \cite{Pieper2006}---an open source package with developmental contributions from multiple agencies including both private and academic institutions.

Related neuroanatomical research concerns the selection of geometric features of the cortex \citep[e.g.][]{Goualher1999} with aims at understanding the functional-anatomical relationship of the human brain. Recent endeavors produce a dense cortical labeling in which every point of the cortex is classified, i.e. a cortical parcellation \citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme time-consumption and tedium associated with manual effort towards the latter, various algorithms have been proposed of which a small subset has been publicly availed to the research community such as the popular software package known as Freesurfer \citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric approach detailed in this work, Freesurfer is primarily a surface-based technique in which the brain structures such as the grey-white matter interface and pial surfaces are processed, analyzed, and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.  Advantages of surface-based approaches include the ability to map processed neuroanatomy to simple geometric primitives such as spheres or planes and the ease of including topological constraints in the analysis workflow.

@@

Our open source segmentation tool, which we have dubbed {\em Atropos},%
\footnote{
Atropos is one of the three Fates from Greek mythology characterized by her dreaded shears used to decide the destiny of each mortal.  Also, consistent with the entomological motif of our ANTs toolkit, {\it Acherontia atropos} is a species of large moth known for the skull-like pattern visible on its thorax.
}
takes advantage of the $n$-tissue paradigm for volumetric image segmentation and harnesses its generalization for template-based cortical parcellation.
Atropos is an EM-based algorithm which contains many of the features of its predecessors for performing
$n$-tissue segmentation including imposition of prior information in the form of MRFs and template-based prior probability maps.  The latter is extended in this work through the formulation of a distance template-based prior used for EM label-propagation and refinement performed in an anatomically meaningful way.

Atropos addresses the bias field issue by incorporating the recently developed N4 bias correction algorithm \citep{Tustison2010} in an adaptive manner.  Coupled with the registration \citep{Avants2010b} and template building strategies \citep{Avants2010} already included in the ANTs toolkit,   
Atropos is a versatile and powerful software tool which touches multiple aspects of our brain processing pipeline from brain extraction \citep{Avants2010a} to grey matter/white matter/cerebrospinal fluid segmentation to label fusion and propagation for achieving a refined cortical parcellation based on both prior information and subject-specific image data.

@@

Bayes' theorem provides a powerful mechanism for making inductive inferences assuming the availability of quantities defining the relevant conditional probabilities, specifically the likelihood and prior probability terms.  A Bayesian paradigm for brain image segmentation utilizes a user-selected observation model defining the likelihood term and one or more prior probability terms whose product is proportional to the posterior probability.  Whereas the likelihood term has been previously defined parametrically (e.g. a Gaussian model) or non-parametrically (e.g. Parzen windowing of the sample histogram), the prior term, as given in the literature, has been formed inclusively either as MRF-based or template-based.  An image segmentation solution in this context is an assignment of one label to each voxel%
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  Following introductory notation, the following sections provide a formal description of these three essential components, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized MRF, template-based prior terms, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}

@@

>>>>

\begin{center}
{\Large \bf An Open Source Framework for Multivariate $n$-Tissue
  Segmentation and Evaluation Using Open Data Sets} \\
\vspace*{0.5cm}
{\normalsize Brian B.~Avants$^{1*}$,  Nicholas J.~Tustison$^2$%
\symbolfootnote[1]{
  The first two authors contributed equally to this work.
}, 
Philip A.~Cook$^1$, Jue Woo$^1$, and James C.~Gee$^1$} \\
\begin{singlespace} 
{\scriptsize  $^1$ Penn Image Computing and Science Laboratory, University of Pennsylvania, Philadelphia, Pennsylvania,  USA.\\
  $^2$ Department of Radiology, University of Virginia, Charlottesville, Virginia, USA}
\end{singlespace}
\end{center}

@@

\begin{abstract} 
Neuroanatomical coordinate systems are essential for the interpretation of 
structural and functional imaging studies.  However, manual delineation of the 
sulco-gyral complex is time consuming and prone to variability given cortical 
complexity.  This work describes an open source, image-based algorithmic approach to 
cortical parcellation which uses training data to propagate structural labelings to 
individual images.  Along with other neuroinformatics tools publicly distributed in our
ANTs (Advanced Normalization Tools)%
\footnote{
http://www.picsl.upenn.edu/ANTs
}
package, {\em Atropos} is an ITK-based multivariate $n$-class
Expectation Maximization (EM) segmentation algorithm which uses either
parametric or non-parametric finite mixture modelings (FMM) of the
class intensities.  It is capable of incorporating prior labelings,
prior probability maps, and/or Markov Random Field (MRF) modeling
assumptions as spatial priors and geodesically propagating
segmentation information across neuroanatomy to ensure a dense
labeling.  Atropos has also been efficiently implemented to handle
large quantities of possible labelings (in the experimental section,
we use up to 68 classes) with a minimal memory footprint.  This work
describes the technical and implementation aspects of Atropos and
evaluates its performance on two different ground-truth datasets.
First, we use the BrainWeb dataset from Montreal Neurological
Institute to evaluate three tissue segmentation performance via (1)
K-means segmentation without use of template data; (2) Markov Random
Field Segmentation with initialization by prior probability maps
derived from a group template; (3) Prior-based Segmentation with use
of spatial prior probability maps derived from a group template.  We
also evaluate Atropos performance by using spatial priors to drive a 68 class EM segmentation
problem defined by the Hammers atlas from University College London
{\bf FIXME proper name}.  These evaluation studies, combined with
illustrative examples that excercise Atropos options, demonstrate both
performance and wide applicability of this new platform-independent 
open source segmentation tool.
%Component (3) corresponds to the most
%realistic clinical case, while component (2) eliminates the confound
%of defining the cortex itself and focuses only on the parcellation
%problem.  Consistent with the open source dissemination of ANTs, we
%make publicly available all scripts and code from the evaluation.
\end{abstract}

@@

\section{Introduction} Neurological and neurodevelopmental research
rely on anatomical and functional correspondence both within and
between subjects.  As image acquisition technology has advanced,
concomitant investment has been made towards image classification
techniques for the cerebrum beginning the appropriation of NASA
satellite image processing software for statistical classification of
head tissues in 2-D MR images \cite{Vannier1985}.  A proliferation of
techniques has ensued with increasing sophistication in both core
methodology and degree of refinement for specific problems.  The
chronology of progress in segmentation may be tracked through both
technical reviews
\citep{Bezdek1993,Pal1993,Clarke1995,Pham2000,Viergever2001,Suri2002,Duncan2004,Balafar2010}
and evaluation studies
\citep[e.g.][]{Cuadra2005,Zaidi2006,Klauschen2009,Boer2010}.

@@

Observations regarding a general trend towards a more integrative view
of neuroanatomical image processing led to the work described in
\cite{Ashburner2005} which is publicly available as SPM---a
large-scale Matlab module in which registration, segmentation, and
bias field correction can be simultaneously modeled within a single
optimization scheme. The roots of this very popular software package
stem back to early seminal work by Karl Friston in which the basis for
statistical parametric mapping was developed \citep{Friston1990} and
subsequently provided to the research community. A similar holistic
approach to brain processing was provided in \cite{Pohl2006} in which
segmentation and registration parameters were optimized simultaneously
while casting the inhomogeneity model parameters of \cite{Wells1996}
as nuisance variables.  Continued work involved recursive parcellation
of the brain volume by considering sub-structures in a hierarchical
manner \citep{Pohl2007}.  An implementation is provided in 3D slicer
\cite{Pieper2006}---an open source package with developmental
contributions from multiple agencies including both private and
academic institutions.  {\bf FIXME references proper names SIENA stuff}
Researchers in aging have also developed tools that have focused on
accurately segmenting the T1 MRI of elderly controls and subjects
suffering from neurodegeneration.  Recent comparative evaluation studies {\bf
  FIXME cite a number of these} have shown that methods such as FAST,
SPM Unified Segmentation, SIENA and others show different performance
characteristics under different evaluation criteria.  That is, no
single method performs best under every measurement, highlighting
the need for segmentation tools that are tunable for different
problems and research goals.

Related neuroanatomical research concerns the selection of geometric
features of the cortex \citep[e.g.][]{Goualher1999} with aims at
understanding the functional-anatomical relationship of the human
brain. Recent endeavors produce a dense cortical labeling in which
every point of the cortex is classified, i.e. a cortical parcellation
\citep{Fischl2004,Heckemann2006,Destrieux2010}.  Given the extreme
time-consumption and tedium associated with manual effort towards the
latter, various algorithms have been proposed of which a small subset
has been publicly availed to the research community such as the
popular software package known as Freesurfer
\citep{Dale1999,Fischl1999,Fischl2004}.  In contrast to the volumetric
approach detailed in this work, Freesurfer is primarily a
surface-based technique in which the brain structures such as the
grey-white matter interface and pial surfaces are processed, analyzed,
and displayed as tessellated surfaces  \cite{Dale1999,Fischl1999}.
Advantages of surface-based approaches include the ability to map
processed neuroanatomy to simple geometric primitives such as spheres
or planes and the ease of including topological constraints in the
analysis workflow.  These types of methods, as well as Klein's
Mindboggle \cite{Klein2003}, would usually follow an initial segmentation by a volumetric method such as Atropos.

@@

Our open source segmentation tool, which we have dubbed {\em
Atropos}, 
\footnote{ Atropos is one of the three Fates from Greek
mythology characterized by her dreaded shears used to decide the
destiny of each mortal.  Also, consistent with the entomological motif
of our ANTs toolkit, {\it Acherontia atropos} is a species of large
moth known for the skull-like pattern visible on its thorax.  }
efficiently and flexibly implements the $n$-tissue paradigm for
volumetric image segmentation.  Atropos allows users to harness its
generalized EM algorithm for standard tissue classification of
the brain into gray matter, white matter and cerebrospinal fluid.
Atropos equally allows incarnations that use EM to simultaneously
maximize the posterior probabilities of many classes, for instance,
when parcellating the brain into hemispheres, cortical regions and
deep brain structures such as amygdala, hippocampus and
thalamus.  Atropos contains features of its predecessors for
performing $n$-tissue segmentation including imposition of prior
information in the form of MRFs and template-based spatial prior
probability maps as well as weighted combinations of these terms.  We
extend the use of spatial priors, in this work, to allow sparse priors
that may then be extended by a geodesic distance-based label propagation
approach.  In short, Atropos seeks to provide a toolbox of
segmentation approaches that may be modified, tuned and refined by the user
for different problems.

We also allow Atropos to address the bias field issue by incorporating
the recently developed N4 bias correction algorithm
\citep{Tustison2010} in an adaptive manner.  Coupled with the
registration \citep{Avants2010b} and template building strategies
\citep{Avants2010} already included in the ANTs toolkit, Atropos is a
versatile and powerful software tool which touches multiple aspects of
our brain processing pipeline from brain extraction
\citep{Avants2010a} to grey matter/white matter/cerebrospinal fluid
segmentation to label fusion and propagation for achieving a refined
cortical parcellation based on both prior information and
subject-specific image data.

@@

Bayes' theorem provides a powerful mechanism for making inductive inferences assuming the availability of quantities defining the relevant conditional probabilities, specifically the likelihood and prior probability terms.  A Bayesian paradigm for brain image segmentation utilizes a user-selected observation model defining the likelihood term and one or more prior probability terms whose product is proportional to the posterior probability.  Whereas the likelihood term has been previously defined parametrically (e.g. a Gaussian model) or non-parametrically (e.g. Parzen windowing of the sample histogram), the prior term, as given in the literature, has been formed inclusively either as MRF-based or template-based.  An image segmentation solution in this context is an assignment of one label to each voxel%
\footnote{
In the classic 3-tissue segmentation case, each voxel in the brain region is assigned a label of `cerebrospinal fluid (csf)', `gray matter (gm)', or `white matter (wm)'. 
}
such that the posterior probability is maximized.  Following
introductory notation, the next sections provide a formal description of these three essential components, viz.
\begin{itemize}
  \item the likelihood or observation model(s),
  \item the prior probability quantities derived from a generalized MRF, template-based prior terms, and a novel distance prior used for label propagation, and 
  \item the optimization framework for maximizing the posterior probability.
\end{itemize}
These key components are common across most segmentation approaches.

@@

{\bf FIXME Prior label image definition?}
